Toward single particle reconstruction without particle picking: Breaking the detection limit

### 2018-08-31 pnas/Towards.tex
updated Appendix A

\TODO{Fill in refs to eqns.} For the experiment in Figure~TKTK, we computed the analytical first three autocorrelations of the volume according to~TKTK. To estimate the coefficients of the volume itself, we solve the LS problem
\begin{equation} \label{eq:LS_cryo}
\min_{\hat{x},\hat{\gamma}} w_1 \vert \mathfrak{a}_x^1 - \hat{\gamma}\mathfrak{a}_{\hat{x}}^1 \vert^2 + w_2\| \mathfrak{a}_x^2 - \hat{\gamma}\mathfrak{a}_{\hat{x}}^2 \|_2^2 + w_3\| \mathfrak{a}_x^3 - \hat{\gamma}\mathfrak{a}_{\hat{x}}^3 \|_F^2, 
\end{equation}  
where the explicit expressions of $\mathfrak{a}_x^1$, $\mathfrak{a}_x^2$ and $\mathfrak{a}_x^3$ are given in~\eqref{eq:ac_1_prolates},~\eqref{eq:ac_2_prolates} and~\eqref{eq:ac_3_prolates}, respectively.  
In the experiments, we set $w_1=w_2=w_3=1$.

The LS problem was solved using Matlab's \texttt{lsqnonlin} solver for nonlinear least-squares problems. The expected $\hat \gamma$ in this experiment is one.

The true volume used in both experiments was the Bovine Pancreatic Trypsin Inhibitor (BPTI) mutant with altered binding loop sequence, whos atomic model is available in the Protein Data Bank (PDB)~\cite{pdb} as 1QLQ. We generated an EM map from this atomic model in UCSF Chimera~\cite{pettersen2004ucsf} at a resolution of 5 \AA, and cropped it to remove zeros at the boundary to obtain a volume of size $31^3$. We also present a reconstruction of TRPV1 in complex with DkTx and RTX, whose EM map is available in the  Electron Microscopy Data Bank (EMDB)~\cite{emdb} as EMD-8117. The original map has size $192^3$, and we downsample it to size $20^3$. To generate the ground truth for our reconstructions, we expanded both volumes as in \eqref{eq:volume_expansion_app} with cutoff $L = 5$. For the TRPV1 reconstruction, the optimizer converged to a point giving relative $\ell_2$ error of $10^{-6}$ in the first three autocorrelations and an error of $10^{-1}$ in the expansion coefficients of the volume. Similarly, \TODO{Fill details for BPTI}. This illustrates the ill-conditioning of the map between the volume and its first three autocorrelations that prevents us from obtaining results from noisy data.

The micrographs for the experiment presented in Figure~\ref{fig:cryo_detection} were generated as follows. We sample rotation matrices from SO(3) uniformly at random using the QR-based algorithm described in QR, and generate the projection of the volume corresponding to that rotation matrix using \verb|cryo_project| in ASPIRE \cite{aspire}. We keep track of the indices at which the upper left corner of a projection can be placed without violating the separation condition, so all projections are separated by at least 30 pixels in each dimension. We continue adding projections to the micrograph until no more projections can be added without violating the separation condition. Then, the LS problem~\eqref{eq:LS_cryo} was solved assuming the spherical harmonic cutoff for the volume is $L=0$, which is sufficient to recover a significant $\widehat\gamma$ in the presence of projections in the micrograph.

>>>>

For the experiment in Figure~\ref{fig:TRPV1_recon}, we computed the analytical first three autocorrelations of the volume as explained in Appendix~\ref{sec:ac_cryo}.
\TODO{Add more ref to figures, if exist}
To estimate the coefficients of the volume itself, we solve the LS problem
\begin{equation} \label{eq:LS_cryo}
\min_{\hat{x},\hat{\gamma}} w_1 \vert \mathfrak{a}_x^1 - \hat{\gamma}\mathfrak{a}_{\hat{x}}^1 \vert^2 + w_2\| \mathfrak{a}_x^2 - \hat{\gamma}\mathfrak{a}_{\hat{x}}^2 \|_2^2 + w_3\| \mathfrak{a}_x^3 - \hat{\gamma}\mathfrak{a}_{\hat{x}}^3 \|_F^2, 
\end{equation}  
where the explicit expressions of $\mathfrak{a}_x^1$, $\mathfrak{a}_x^2$ and $\mathfrak{a}_x^3$ are given in~\eqref{eq:ac_1_prolates},~\eqref{eq:ac_2_prolates} and~\eqref{eq:ac_3_prolates}, respectively.  
In the experiments, we set $w_1=w_2=w_3=1$. 
The LS problem was solved using Matlab's \texttt{lsqnonlin} solver for nonlinear LS problems. The expected $\hat \gamma$ in this experiment is one.

The true volume used in both experiments was the Bovine Pancreatic Trypsin Inhibitor (BPTI) mutant with altered binding loop sequence, whos atomic model is available in the Protein Data Bank (PDB)~\cite{pdb} as 1QLQ. We generated an EM map from this atomic model in UCSF Chimera~\cite{pettersen2004ucsf} at a resolution of 5 \AA, and cropped it to remove zeros at the boundary to obtain a volume of size $31^3$. We also present a reconstruction of TRPV1 in complex with DkTx and RTX, whose EM map is available in the  Electron Microscopy Data Bank (EMDB)~\cite{emdb} as EMD-8117. The original map has size $192^3$, and we downsample it to size $20^3$. To generate the ground truth for our reconstructions, we expanded both volumes as in \eqref{eq:volume_expansion_app} with cutoff $L = 5$. For the TRPV1 reconstruction, the optimizer converged to a point giving relative $\ell_2$ error of $10^{-6}$ in the first three autocorrelations and an error of $10^{-1}$ in the expansion coefficients of the volume. Similarly, \TODO{Fill details for BPTI}. This illustrates the ill-conditioning of the map between the volume and its first three autocorrelations that prevents us from obtaining results from noisy data. \TODO{Tamir: This entire paragraph will be changed a bit once we will have the final figures}

The micrographs for the experiment presented in Figure~\ref{fig:cryo_detection} were generated as follows. We sample rotation matrices from SO(3) uniformly at random using the QR-based algorithm described in~\cite{stewart1980efficient}, and generate the projection of the volume corresponding to that rotation matrix using \verb|cryo_project| in ASPIRE \cite{aspire}. We keep track of the indices at which the upper left corner of a projection can be placed without violating the separation condition, so all projections are separated by at least 30 pixels in each dimension. We continue adding projections to the micrograph until no more projections can be added without violating the separation condition. Then, the LS problem~\eqref{eq:LS_cryo} was solved assuming the spherical harmonic cutoff for the volume is $L=0$, which is sufficient to recover a significant $\widehat\gamma$ in the presence of projections in the micrograph.
